---
title: "`MLML2R`: an R package for maximum likelihood estimates of DNA methylation and hydroxymethylation"
author: Samara F. Kiihl, Maria Tellez-Plaza
output:
  html_document:
    keep_md: yes
    toc: true
    number_sections: true
---



# Introduction


This document presents an example of the usage of the `MLML2R` package for R.

Install the R package using the following commands on the R console:

```{r,echo=TRUE,eval=FALSE}
install.packages("devtools")
devtools::install_github("samarafk/MLML2R")
library(MLML2R)
```


The function `MLML` provides maximum likelihood estimates (MLE) for 5-hmC and 5-mC levels using data from any combination of two of the methods: BS-seq, TAB-seq or oxBS-seq, or combination of all the three methods.

The algorithm implemented in the `MLML` function is based on the Expectation-Maximization (EM) algorithm proposed by [Qu *et al.* (2013)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3789553/). In addition, our implementation is optimized, since we derived the exact constrained MLE for 5-mC or 5-hmC levels, and the iterative EM algorithm is not needed. Our improved formulation can, thus, decrease analytic processing time and computational burden, common bottlenecks when processing single-base profiling data from thousands of samples.

Furthermore, our routine is flexible and can be used with both next generation sequencing and Infinium Methylation microarray data in the R-statistical language.



# BS+oxBS+TAB data

```{r}
library(MLML2R)

# via EM algortihm
results_em <- MLML(T = MethylatedBS_sim2 , U = UnMethylatedBS_sim2,
                   L = UnMethylatedOxBS_sim2, M = MethylatedOxBS_sim2,
                   G = UnMethylatedTAB_sim2, H = MethylatedTAB_sim2,tol=0.00001,iterative = TRUE)

# via Lagrange multiplier approximation
results_lag <- MLML(T = MethylatedBS_sim2 , U = UnMethylatedBS_sim2,
                   L = UnMethylatedOxBS_sim2, M = MethylatedOxBS_sim2,
                   G = UnMethylatedTAB_sim2, H = MethylatedTAB_sim2)
```

```{r,echo=FALSE,fig.align='center',fig.height=8,fig.width=8}
plot(results_em$hmC,results_lag$hmC)
```

```{r}
all.equal(results_em$hmC,results_lag$hmC)

all.equal(results_em$C,results_lag$C)
```

```{r}

library(microbenchmark)
mbm = microbenchmark(
   lagrange = MLML(T = MethylatedBS_sim2 , U = UnMethylatedBS_sim2,
                                                             L = UnMethylatedOxBS_sim2, M = MethylatedOxBS_sim2,
                                                            G = UnMethylatedTAB_sim2, H = MethylatedTAB_sim2),
   EM = MLML(T = MethylatedBS_sim2 , U = UnMethylatedBS_sim2,
                            L = UnMethylatedOxBS_sim2, M = MethylatedOxBS_sim2,
                            G = UnMethylatedTAB_sim2, H = MethylatedTAB_sim2,tol=0.0001,iterative = TRUE),
   times=10)
mbm



T = MethylatedBS_sim2
U = UnMethylatedBS_sim2
L = UnMethylatedOxBS_sim2
M = MethylatedOxBS_sim2
G = UnMethylatedTAB_sim2
H = MethylatedTAB_sim2


results_oxBS_TAB_BS_exact <- list()

results_oxBS_TAB_BS_exact$mC <- M/(U+H+M)
results_oxBS_TAB_BS_exact$hmC <- H/(U+H+M)
results_oxBS_TAB_BS_exact$C <- U/(U+H+M)


```{r,echo=FALSE,fig.align='center',fig.height=3,fig.width=9}

par(mfrow =c(1,3))

plot(density(results_oxBS_TAB_BS_exact$hmC[,1]),main= "5-hmC using exact unconstrained",xlab=" ",ylim=c(0,15),xlim=c(0,1))
lines(density(results_oxBS_TAB_BS_exact$hmC[,2]),col=2)


plot(density(results_oxBS_TAB_BS_exact$mC[,1]),main= "5-mC using exact unconstrained",xlab=" ",ylim=c(0,5),xlim=c(0,1))
lines(density(results_oxBS_TAB_BS_exact$mC[,2]),col=2)


plot(density(results_oxBS_TAB_BS_exact$C[,1]),main= "5-C using exact unconstrained",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_oxBS_TAB_BS_exact$C[,2]),col=2)
```



```{r,echo=FALSE,fig.align='center',fig.height=3,fig.width=9}

par(mfrow =c(1,3))

plot(density(results_lag$hmC[,1]),main= "5-hmC using Lagrange",xlab=" ",xlim=c(0,1),ylim=c(0,15))
lines(density(results_lag$hmC[,2]),col=2)

plot(density(results_lag$mC[,1]),main= "5-mC using Lagrange",xlab=" ",xlim=c(0,1),ylim=c(0,5))
lines(density(results_lag$mC[,2]),col=2)

plot(density(results_lag$C[,1]),main= "5-C using Lagrange",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_lag$C[,2]),col=2)
```


```{r,echo=FALSE,fig.align='center',fig.height=3,fig.width=9}

par(mfrow =c(1,3))

plot(density(results_em$hmC[,1]),main= "5-hmC using EM",xlab=" ",xlim=c(0,1),ylim=c(0,15))
lines(density(results_em$hmC[,2]),col=2)

plot(density(results_em$mC[,1]),main= "5-mC using EM",xlab=" ",xlim=c(0,1),ylim=c(0,5))
lines(density(results_em$mC[,2]),col=2)

plot(density(results_em$C[,1]),main= "5-C using EM",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_em$C[,2]),col=2)
```


```{r,echo=FALSE,fig.align='center'}
all.equal(results_em$hmC,results_oxBS_TAB_BS_exact$hmC)
```

