---
title: "Example data for `MLML2R` package"
output:
  html_document:
    keep_md: yes
    toc: true
    number_sections: true
---


# Getting publicly available data


We will use the dataset from [Field *et al.* (2015)](https://doi.org/10.1371/journal.pone.0118202), which consists of eight DNA samples from the same DNA source treated with oxBS-BS and hybridized to the Infinium 450K array.

The steps shown in this section follows the [vignette](https://kasperdanielhansen.github.io/genbioconductor/html/minfi.html) from `minfi` package.

We start with the steps to get the raw data from the GEO repository.
The dataset from [Field *et al.* (2015)](https://doi.org/10.1371/journal.pone.0118202) is available at GEO accession [GSE63179](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63179).

The sample was divided into four BS and four oxBS replicates.
 	
Platform used: GPL16304	Illumina HumanMethylation450 BeadChip [UBC enhanced annotation v1.0]

Samples:

* GSM1543269	brain-BS-1
* GSM1543270	brain-oxBS-3
* GSM1543271	brain-BS-2
* GSM1543272	brain-oxBS-4
* GSM1543273	brain-BS-3
* GSM1543274	brain-BS-4
* GSM1543275	brain-oxBS-1
* GSM1543276	brain-oxBS-2

This example has the following dependencies:

```{r,warning=FALSE,message=FALSE}
library(minfi)
library(GEOquery)
```

Use the following commands to install these packages in R:
```{r,echo=TRUE,eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("minfi", "GEOquery"))
```


```{r,echo=FALSE}
if(! file.exists("GSE63179/GSE63179_RAW.tar"))
{
getGEOSuppFiles("GSE63179")
untar("GSE63179/GSE63179_RAW.tar", exdir = "GSE63179/idat")
head(list.files("GSE63179/idat", pattern = "idat"))
}
```


```{r,echo=TRUE,eval=FALSE}
getGEOSuppFiles("GSE63179")
untar("GSE63179/GSE63179_RAW.tar", exdir = "GSE63179/idat")
head(list.files("GSE63179/idat", pattern = "idat"))
```


Decompress the compressed IDAT files:

```{r,message=FALSE,warning=FALSE}
idatFiles <- list.files("GSE63179/idat", pattern = "idat.gz$", full = TRUE)
sapply(idatFiles, gunzip, overwrite = TRUE)
```


Now we read the IDAT files in the directory:

```{r}
rgSet <- read.metharray.exp("GSE63179/idat")
rgSet
```

```{r}
pData(rgSet)
```

```{r}
sampleNames(rgSet)
```

The file names consists of a GEO identifier (the GSM part) followed by a standard IDAT naming convention with a 10 digit number which is an array identifier followed by an identifier of the form R01C01. This is because each array actually allows for the hybridization of 12 samples in a 6x2 arrangement. The 9373551079_R01C01 means row 1 and column 1 on chip 9373551079. 

We need to identify the samples from different methods: BS-conversion, oxBS-conversion.

```{r,echo=FALSE}
if (!file.exists("pD.rds"))
{
geoMat <- getGEO("GSE63179")
pD.all <- pData(geoMat[[1]])
pD <- pD.all[, c("title", "geo_accession", "characteristics_ch1.1", "characteristics_ch1.2","characteristics_ch1.3")]
save(pD,file="pD.rds")
}
```


```{r,eval=FALSE}
geoMat <- getGEO("GSE63179")
pD.all <- pData(geoMat[[1]])
pD <- pD.all[, c("title", "geo_accession", "characteristics_ch1.1", "characteristics_ch1.2","characteristics_ch1.3")]
pD
```

```{r,echo=FALSE}
load("pD.rds")
pD
```

```{r}
names(pD)[c(3,4,5)] <- c("gender", "age","method")
pD$gender <- sub("^gender: ", "", pD$gender)
pD$age <- sub("^age: ", "", pD$age)
pD$method <- sub("^bisulfite_proc: ","",pD$method)
```

We now need to merge this pheno data into the methylation data. The following are commands to make sure we have the same row identifier in both datasets before merging.

```{r}
sampleNames(rgSet) <- sapply(sampleNames(rgSet),function(x) strsplit(x,"_")[[1]][1])
rownames(pD) <- pD$geo_accession
pD <- pD[sampleNames(rgSet),]
pData(rgSet) <- as(pD,"DataFrame")
rgSet
```


# Preprocessing

We refer the reader to the `minfi` package tutorials for more preprocessing options.


We need to install the required package bellow:
```{r,eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("IlluminaHumanMethylation450kmanifest")
```


First, we removed probes with detection p-value <0.01 in any of the 8 arrays. The function `detectionP` identifies failed positions defined as both the methylated and unmethylated channel reporting background signal levels.

```{r,message=FALSE}
detP <- detectionP(rgSet)
failed <- detP >0.01
## Keep probes which failed in at most maxFail arrays (0 = the probe passed in all arrays)
maxFail<- 0
keep_probes <- rowSums(failed) <= maxFail
```

We kept $`r round(mean(keep_probes)*100,0)`\%$ of the probes according to this criterion.


The `rgSet` object is a class called `RGChannelSet` which represents two color data with a green and a red channel. We will use, as input in the `MLML` funcion, a `MethylSet`, which contains the methylated and unmethylated signals. The most basic way to construct a `MethylSet` is to using the function `preprocessRaw` which uses the array design to match up the different probes and color channels to construct the methylated and unmethylated signals. Here we will use the `preprocessNoob` function, which does the preprocessing and returns a `MethylSet`.


Arrays were then normalized using the Noob/ssNoob preprocessing method for Infinium methylation microarrays.


From a `MethylSet` it is easy to compute Beta values, defined as:

Beta = Meth / (Meth + Unmeth + c)

The c constant is chosen to avoid dividing with small values. Illumina uses a default of c=100. The function `getBeta` from `minfi` package can be used to obtain the Beta values.


```{r,echo=TRUE,message=FALSE}
MSet.noob<- preprocessNoob(rgSet[keep_probes,])
```


Prepare de input data:


```{r,echo=TRUE}
MethylatedBS <- getMeth(MSet.noob)[,c(1,3,5,6)]
UnMethylatedBS <- getUnmeth(MSet.noob)[,c(1,3,5,6)]

MethylatedOxBS <- getMeth(MSet.noob)[,c(7,8,2,4)]
UnMethylatedOxBS <- getUnmeth(MSet.noob)[,c(7,8,2,4)]
```


```{r,echo=FALSE,eval=FALSE}
# not used in the package
# Only a small sample of CpGs and 2 replicates will be used in the example data.

set.seed(112017)

a <- sample(1:dim(MethylatedBS)[1],100,replace=FALSE)

MethylatedBS <- MethylatedBS[a,1:2]
UnMethylatedBS <- UnMethylatedBS[a,1:2]
MethylatedOxBS <- MethylatedOxBS[a,1:2]
UnMethylatedOxBS <- UnMethylatedOxBS[a,1:2]

colnames(MethylatedBS) <- c("sample1","sample2")
colnames(UnMethylatedBS) <- c("sample1","sample2")
colnames(MethylatedOxBS) <- c("sample1","sample2")
colnames(UnMethylatedOxBS) <- c("sample1","sample2")

save(MethylatedBS,file="MethylatedBS.rda")
save(UnMethylatedBS,file="UnMethylatedBS.rda")
save(MethylatedOxBS,file="MethylatedOxBS.rda")
save(UnMethylatedOxBS,file="UnMethylatedOxBS.rda")
```


# This was included in the package as example:

```{r}
set.seed(112017)

a <- sample(1:dim(MethylatedBS)[1],100,replace=FALSE)

MethylatedBS <- MethylatedBS[a,1:2]
UnMethylatedBS <- UnMethylatedBS[a,1:2]
MethylatedOxBS <- MethylatedOxBS[a,1:2]
UnMethylatedOxBS <- UnMethylatedOxBS[a,1:2]

colnames(MethylatedBS) <- c("sample1","sample2")
colnames(UnMethylatedBS) <- c("sample1","sample2")
colnames(MethylatedOxBS) <- c("sample1","sample2")
colnames(UnMethylatedOxBS) <- c("sample1","sample2")

N_BS <- round(MethylatedBS+UnMethylatedBS)

N_OxBS <- round(MethylatedOxBS+UnMethylatedOxBS)

N_TAB <- pmax(N_BS,N_OxBS)

p_m=.3
p_h=0.2
p_u=.5

true_parameters_sim <- data.frame(p_m=.3,p_h=.2,p_u=.5)
save(true_parameters_sim,file="true_parameters_sim.rda")

set.seed(2017)
MethylatedBS_sim <- apply(N_BS, c(1,2), function(x) rbinom(n=1, size=x, prob=(p_m+p_h)))
                          
UnMethylatedBS_sim <- N_BS - MethylatedBS_sim

MethylatedOxBS_sim <- apply(N_OxBS, c(1,2), function(x) rbinom(n=1, size=x, prob=p_m))
                            
UnMethylatedOxBS_sim <- N_OxBS - MethylatedOxBS_sim

MethylatedTAB_sim <- apply(N_TAB, c(1,2), function(x) rbinom(n=1, size=x, prob=p_h))
                           
UnMethylatedTAB_sim <- N_TAB - MethylatedTAB_sim


save(MethylatedBS_sim,file="MethylatedBS_sim.rda")
save(UnMethylatedBS_sim,file="UnMethylatedBS_sim.rda")
save(MethylatedOxBS_sim,file="MethylatedOxBS_sim.rda")
save(UnMethylatedOxBS_sim,file="UnMethylatedOxBS_sim.rda")
save(MethylatedTAB_sim,file="MethylatedTAB_sim.rda")
save(UnMethylatedTAB_sim,file="UnMethylatedTAB_sim.rda")

```



# Second Example Dataset

```{r}
MethylatedBS <- getMeth(MSet.noob)[,c(1,3,5,6)]
UnMethylatedBS <- getUnmeth(MSet.noob)[,c(1,3,5,6)]

MethylatedOxBS <- getMeth(MSet.noob)[,c(7,8,2,4)]
UnMethylatedOxBS <- getUnmeth(MSet.noob)[,c(7,8,2,4)]

colnames(MethylatedBS) <- c("sample1","sample2","sample3","sample4")
colnames(UnMethylatedBS) <- c("sample1","sample2","sample3","sample4")
colnames(MethylatedOxBS) <- c("sample1","sample2","sample3","sample4")
colnames(UnMethylatedOxBS) <- c("sample1","sample2","sample3","sample4")

CpG <- rownames(MethylatedBS)

N_BS <- round(MethylatedBS+UnMethylatedBS)

N_OxBS <- round(MethylatedOxBS+UnMethylatedOxBS)

N_TAB <- pmax(N_BS,N_OxBS)

# Use estimates from dataset as base for the true parameters in the simulation
library(MLML2R)
results_exact <- MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, M = MethylatedOxBS)


set.seed(2017)

temp1 <- data.frame(n=as.vector(N_BS),p_m=c(results_exact$mC[,1],results_exact$mC[,1],results_exact$mC[,1],results_exact$mC[,1]),p_h=c(results_exact$hmC[,1],results_exact$hmC[,1],results_exact$hmC[,1],results_exact$hmC[,1]))

MethylatedBS_temp <- c()
for (i in 1:dim(temp1)[1])
{
  MethylatedBS_temp[i] <- rbinom(n=1, size=temp1$n[i], prob=(temp1$p_m[i]+temp1$p_h[i]))
}

UnMethylatedBS_sim2 <- matrix(N_BS - MethylatedBS_temp,ncol=4)
MethylatedBS_sim2 <- matrix(MethylatedBS_temp,ncol=4)

temp1 <- data.frame(n=as.vector(N_OxBS),p_m=c(results_exact$mC[,1],results_exact$mC[,1],results_exact$mC[,1],results_exact$mC[,1]),p_h=c(results_exact$hmC[,1],results_exact$hmC[,1],results_exact$hmC[,1],results_exact$hmC[,1]))

MethylatedOxBS_temp <- c()
for (i in 1:dim(temp1)[1])
{
  MethylatedOxBS_temp[i] <- rbinom(n=1, size=temp1$n[i], prob=temp1$p_m[i])
}

UnMethylatedOxBS_sim2 <- matrix(N_OxBS - MethylatedOxBS_temp,ncol=4)
MethylatedOxBS_sim2 <- matrix(MethylatedOxBS_temp,ncol=4)


temp1 <- data.frame(n=as.vector(N_TAB),p_m=c(results_exact$mC[,1],results_exact$mC[,1],results_exact$mC[,1],results_exact$mC[,1]),p_h=c(results_exact$hmC[,1],results_exact$hmC[,1],results_exact$hmC[,1],results_exact$hmC[,1]))

MethylatedTAB_temp <- c()
for (i in 1:dim(temp1)[1])
{
  MethylatedTAB_temp[i] <- rbinom(n=1, size=temp1$n[i], prob=temp1$p_h[i])
}


UnMethylatedTAB_sim2 <- matrix(N_TAB - MethylatedTAB_temp,ncol=4)
MethylatedTAB_sim2 <- matrix(MethylatedTAB_temp,ncol=4)


set.seed(112017)

a <- sample(1:dim(MethylatedBS)[1],1000,replace=FALSE)

MethylatedBS_sim2 <- MethylatedBS_sim2[a,]
UnMethylatedBS_sim2 <- UnMethylatedBS_sim2[a,]
MethylatedOxBS_sim2 <- MethylatedOxBS_sim2[a,]
UnMethylatedOxBS_sim2 <- UnMethylatedOxBS_sim2[a,]
MethylatedTAB_sim2 <- MethylatedTAB_sim2[a,]
UnMethylatedTAB_sim2 <- UnMethylatedTAB_sim2[a,]

colnames(MethylatedBS_sim2) <- c("sample1","sample2","sample3","sample4")
colnames(UnMethylatedBS_sim2) <- c("sample1","sample2","sample3","sample4")
colnames(MethylatedOxBS_sim2) <- c("sample1","sample2","sample3","sample4")
colnames(UnMethylatedOxBS_sim2) <- c("sample1","sample2","sample3","sample4")

rownames(MethylatedBS_sim2) <- CpG[a]
rownames(UnMethylatedBS_sim2) <- CpG[a]
rownames(MethylatedOxBS_sim2) <- CpG[a]
rownames(UnMethylatedOxBS_sim2) <- CpG[a]
rownames(MethylatedTAB_sim2) <- CpG[a]
rownames(UnMethylatedTAB_sim2) <- CpG[a]


save(MethylatedBS_sim2,file="MethylatedBS_sim2.rda")
save(UnMethylatedBS_sim2,file="UnMethylatedBS_sim2.rda")
save(MethylatedOxBS_sim2,file="MethylatedOxBS_sim2.rda")
save(UnMethylatedOxBS_sim2,file="UnMethylatedOxBS_sim2.rda")
save(MethylatedTAB_sim2,file="MethylatedTAB_sim2.rda")
save(UnMethylatedTAB_sim2,file="UnMethylatedTAB_sim2.rda")


true_parameters_sim2 <- data.frame(p_m=results_exact$mC[,1],p_h=results_exact$hmC[,1])
true_parameters_sim2$p_u <- 1-true_parameters_sim2$p_m-true_parameters_sim2$p_h
true_parameters_sim2 <- true_parameters_sim2[a,]
save(true_parameters_sim2,file="true_parameters_sim2.rda")

```
