---
title: "`MLML2R`: an R package for maximum likelihood estimates of DNA methylation and hydroxymethylation"
author: Samara F. Kiihl, Maria Tellez-Plaza
output:
  html_document:
    keep_md: yes
    toc: true
    number_sections: true
---


# Introduction


This document presents an example of the usage of the `MLML2R` package for R.

Install the R package using the following commands on the R console:

```{r,echo=TRUE,eval=FALSE}
install.packages("devtools")
devtools::install_github("samarafk/MLML2R")
library(MLML2R)
```


Proposed analyses of single-base profiling of either 5-hmC or 5-mC require combining data obtained using bisulfite conversion, oxidative bisulfite conversion or Tet-Assisted bisulfite conversion methods, but doing so naively produces inconsistent estimates of 5-mC or 5-hmC level [(Qu *et al.*, 2013)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3789553/). 


The function `MLML` provides maximum likelihood estimates (MLE) for 5-hmC and 5-mC levels using data from any combination of two of the methods: BS-seq, TAB-seq or oxBS-seq. The function also provides MLE when combining these three methods.

The algorithm implemented in the `MLML` function is based on the Expectation-Maximization (EM) algorithm proposed by [Qu *et al.* (2013)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3789553/). In addition, when only two methods are combined, our implementation is optimized, since we derived the exact MLE for 5-mC or 5-hmC levels, and the iterative EM algorithm is not needed. Our improved formulation can, thus, decrease analytic processing time and computational burden, common bottlenecks when processing single-base profiling data from thousands of samples.

Furthermore, our routine is flexible and can be used with both next generation sequencing and Infinium Methylation microarray data in the R-statistical language.



# Preparing dataset

We will use the dataset from [Field *et al.* (2015)](https://doi.org/10.1371/journal.pone.0118202), which consists of eight DNA samples from the same DNA source treated with oxBS-BS and hybridized to the Infinium 450K array.

The steps shown in this section follows the [vignette](https://kasperdanielhansen.github.io/genbioconductor/html/minfi.html) from `minfi` package.

## Getting publicly available data

We start with the steps to get the raw data from the GEO repository.
The dataset from [Field *et al.* (2015)](https://doi.org/10.1371/journal.pone.0118202) is available at GEO accession [GSE63179](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63179).

The sample was divided into four BS and four oxBS replicates.
 	
Platform used: GPL16304	Illumina HumanMethylation450 BeadChip [UBC enhanced annotation v1.0]

Samples:

* GSM1543269	brain-BS-1
* GSM1543270	brain-oxBS-3
* GSM1543271	brain-BS-2
* GSM1543272	brain-oxBS-4
* GSM1543273	brain-BS-3
* GSM1543274	brain-BS-4
* GSM1543275	brain-oxBS-1
* GSM1543276	brain-oxBS-2



This example has the following dependencies:

```{r,warning=FALSE,message=FALSE}
library(minfi)
library(GEOquery)
```

Use the following commands to install these packages in R:
```{r,echo=TRUE,eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("minfi", "GEOquery"))
```



```{r,echo=FALSE}
if(! file.exists("GSE63179/GSE63179_RAW.tar"))
{
getGEOSuppFiles("GSE63179")
untar("GSE63179/GSE63179_RAW.tar", exdir = "GSE63179/idat")
head(list.files("GSE63179/idat", pattern = "idat"))
}
```


```{r,echo=TRUE,eval=FALSE}
getGEOSuppFiles("GSE63179")
untar("GSE63179/GSE63179_RAW.tar", exdir = "GSE63179/idat")
head(list.files("GSE63179/idat", pattern = "idat"))
```


Decompress the compressed IDAT files:

```{r,message=FALSE,warning=FALSE}
idatFiles <- list.files("GSE63179/idat", pattern = "idat.gz$", full = TRUE)
sapply(idatFiles, gunzip, overwrite = TRUE)
```


Now we read the IDAT files in the directory:

```{r}
rgSet <- read.metharray.exp("GSE63179/idat")
rgSet
```

```{r}
pData(rgSet)
```

```{r}
sampleNames(rgSet)
```

The file names consists of a GEO identifier (the GSM part) followed by a standard IDAT naming convention with a 10 digit number which is an array identifier followed by an identifier of the form R01C01. This is because each array actually allows for the hybridization of 12 samples in a 6x2 arrangement. The 9373551079_R01C01 means row 1 and column 1 on chip 9373551079. 

We need to identify the samples from different methods: BS-conversion, oxBS-conversion.

```{r,echo=FALSE}
if (!file.exists("pD.rds"))
{
geoMat <- getGEO("GSE63179")
pD.all <- pData(geoMat[[1]])
pD <- pD.all[, c("title", "geo_accession", "characteristics_ch1.1", "characteristics_ch1.2","characteristics_ch1.3")]
save(pD,file="pD.rds")
}
```


```{r,eval=FALSE}
geoMat <- getGEO("GSE63179")
pD.all <- pData(geoMat[[1]])
pD <- pD.all[, c("title", "geo_accession", "characteristics_ch1.1", "characteristics_ch1.2","characteristics_ch1.3")]
pD
```

```{r,echo=FALSE}
load("pD.rds")
pD
```

```{r}
names(pD)[c(3,4,5)] <- c("gender", "age","method")
pD$gender <- sub("^gender: ", "", pD$gender)
pD$age <- sub("^age: ", "", pD$age)
pD$method <- sub("^bisulfite_proc: ","",pD$method)
```

We now need to merge this pheno data into the methylation data. The following are commands to make sure we have the same row identifier in both datasets before merging.

```{r}
sampleNames(rgSet) <- sapply(sampleNames(rgSet),function(x) strsplit(x,"_")[[1]][1])
rownames(pD) <- pD$geo_accession
pD <- pD[sampleNames(rgSet),]
pData(rgSet) <- as(pD,"DataFrame")
rgSet

save(rgSet,file="rgSet.rds")
```


## Preprocessing

We refer the reader to the `minfi` package tutorials for more preprocessing options.


We need to install the required package bellow:
```{r,eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("IlluminaHumanMethylation450kmanifest")
```


First, we removed probes with detection p-value <0.01 in any of the 8 arrays. The function `detectionP` identifies failed positions defined as both the methylated and unmethylated channel reporting background signal levels.

```{r,message=FALSE}
detP <- detectionP(rgSet)
failed <- detP >0.01
## Keep probes which failed in at most maxFail arrays (0 = the probe passed in all arrays)
maxFail<- 0
keep_probes <- rowSums(failed) <= maxFail
```

We kept $`r round(mean(keep_probes)*100,0)`\%$ of the probes according to this criterion.


The `rgSet` object is a class called `RGChannelSet` which represents two color data with a green and a red channel. We will use, as input in the `MLML` funcion, a `MethylSet`, which contains the methylated and unmethylated signals. The most basic way to construct a `MethylSet` is to using the function `preprocessRaw` which uses the array design to match up the different probes and color channels to construct the methylated and unmethylated signals. Here we will use the `preprocessNoob` function, which does the preprocessing and returns a `MethylSet`.


```{r,message=FALSE,eval=FALSE,echo=FALSE}
MSet <- preprocessRaw(rgSet)
MSet <- MSet[keep_probes,] # keep good probes
```

Arrays were then normalized using the Noob/ssNoob preprocessing method for Infinium methylation microarrays.


From a `MethylSet` it is easy to compute Beta values, defined as:

Beta = Meth / (Meth + Unmeth + c)

The c constant is chosen to avoid dividing with small values. Illumina uses a default of c=100. The function `getBeta` from `minfi` package can be used to obtain the Beta values.


```{r,echo=TRUE,message=FALSE}
MSet.noob<- preprocessNoob(rgSet[keep_probes,])
densityPlot(MSet.noob, sampGroups= pData(rgSet)$method,
main= sprintf('Beta values for filtered probes (n= %s)', nrow(MSet.noob)))
```


```{r,echo=FALSE,eval=FALSE}
MSet.swan<- preprocessSWAN(rgSet, mSet = MSet)
densityPlot(MSet.swan, sampGroups= pData(rgSet)$method,
main= sprintf('Beta values for filtered probes (n= %s)', nrow(MSet.swan)))
```

# Using the `MLML2R` package

After all the preprocessing procedures, we now can use the `MLML2R` package to obtain the maximum likelihood estimates for the 5-hmC and 5-mC levels.

Install the R package using the following commands on the R console:


```{r,eval=FALSE}
install.packages("devtools")
devtools::install_github("samarafk/MLML2R")
```


Prepare de input data:
```{r,echo=FALSE,eval=FALSE}
MethylatedBS <- getMeth(MSet.swan)[,c(1,3,5,6)]
UnMethylatedBS <- getUnmeth(MSet.swan)[,c(1,3,5,6)]

MethylatedOxBS <- getMeth(MSet.swan)[,c(7,8,2,4)]
UnMethylatedOxBS <- getUnmeth(MSet.swan)[,c(7,8,2,4)]
```

```{r,echo=TRUE}
MethylatedBS <- getMeth(MSet.noob)[,c(1,3,5,6)]
UnMethylatedBS <- getUnmeth(MSet.noob)[,c(1,3,5,6)]

MethylatedOxBS <- getMeth(MSet.noob)[,c(7,8,2,4)]
UnMethylatedOxBS <- getUnmeth(MSet.noob)[,c(7,8,2,4)]
```


```{r,echo=FALSE}
save(MethylatedBS,UnMethylatedBS,MethylatedOxBS,UnMethylatedOxBS, file="MethylationInputData.rds")
```


Getting the MLE estimates using EM-algorithm:

```{r}
library(MLML2R)
results_em <- MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, M = MethylatedOxBS,tol=0.0001)
save(results_em,file="results_em.rds")
```


Plot of the results (we have 4 replicates)

```{r,fig.align='center',fig.height=3,fig.width=9}
par(mfrow =c(1,3)) 

plot(density(results_em$hmC[,1]),main= "5-hmC using MLML",xlab=" ",xlim=c(0,1))
lines(density(results_em$hmC[,2]),col=2)
lines(density(results_em$hmC[,3]),col=3)
lines(density(results_em$hmC[,4]),col=4)

plot(density(results_em$mC[,1]),main= "5-mC using MLML",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_em$mC[,2]),col=2)
lines(density(results_em$mC[,3]),col=3)
lines(density(results_em$mC[,4]),col=4)

plot(density(results_em$C[,1]),main= "5-C using MLML",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_em$C[,2]),col=2)
lines(density(results_em$C[,3]),col=3)
lines(density(results_em$C[,4]),col=4)
```


Getting the constrained exact MLE estimates:

```{r}
library(MLML2R)
results_exact <- MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, M = MethylatedOxBS,exact=TRUE)
save(results_exact,file="results_exact.rds")
```


Plot of the results (we have 4 replicates)

```{r,fig.align='center',fig.height=3,fig.width=9}
par(mfrow =c(1,3)) 

plot(density(results_exact$hmC[,1]),main= "5-hmC using MLML",xlab=" ",xlim=c(0,1))
lines(density(results_exact$hmC[,2]),col=2)
lines(density(results_exact$hmC[,3]),col=3)
lines(density(results_exact$hmC[,4]),col=4)

plot(density(results_exact$mC[,1]),main= "5-mC using MLML",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_exact$mC[,2]),col=2)
lines(density(results_exact$mC[,3]),col=3)
lines(density(results_exact$mC[,4]),col=4)

plot(density(results_exact$C[,1]),main= "5-C using MLML",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(results_exact$C[,2]),col=2)
lines(density(results_exact$C[,3]),col=3)
lines(density(results_exact$C[,4]),col=4)
```

# Other methods to obtain the estimates

## Naive estimates

The naive approach to obtain 5-hmC levels is $\beta_{BS} -  \beta_{OxBS}$. This approach results in negative values for the 5-hmC levels.

```{r,echo=TRUE}
beta_BS <- getBeta(MSet.noob)[,c(1,3,5,6)]
beta_OxBS <- getBeta(MSet.noob)[,c(7,8,2,4)]
hmC_naive <- beta_BS-beta_OxBS
C_naive <- 1-beta_BS
mC_naive <- beta_OxBS
```


```{r,fig.align='center',fig.height=3,fig.width=9}
par(mfrow =c(1,3)) 

plot(density(hmC_naive[,1]),main= "5-hmC using naive",ylim=c(0,8),xlab=" ",xlim=c(-1,1))
lines(density(hmC_naive[,2]),col=2)
lines(density(hmC_naive[,3]),col=3)
lines(density(hmC_naive[,4]),col=4)

plot(density(mC_naive[,1]),main= "5-mC using naive",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(mC_naive[,2]),col=2)
lines(density(mC_naive[,3]),col=3)
lines(density(mC_naive[,4]),col=4)

plot(density(C_naive[,1]),main= "5-C using naive",ylim=c(0,5),xlab=" ",xlim=c(0,1))
lines(density(C_naive[,2]),col=2)
lines(density(C_naive[,3]),col=3)
lines(density(C_naive[,4]),col=4)
```


## `OxyBS` estimates

For the specific case where only ox-BS and BS data are available, `OxyBS` package from [Houseman *et al.* (2016)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4978924/) can be use to obtain estimates.



```{r,echo=TRUE,eval=FALSE}
library(OxyBS)

# Methylated signals from the BS and oxBS arrays
methBS <- MethylatedBS
methOxBS <- MethylatedOxBS
# Unmethylated signals from the BS and oxBS arrays
unmethBS <- UnMethylatedBS
unmethOxBS <- UnMethylatedOxBS

# Calculate Total Signals
signalBS <- methBS+unmethBS
signalOxBS <- methOxBS+unmethOxBS

# Calculate Beta Values
betaBS <- methBS/signalBS
betaOxBS <- methOxBS/signalOxBS

####################################################
# 4. Apply fitOxBS function to preprocessed values
####################################################

# Select the number of CpGs and Subjects to which the method will be applied 
nCpGs <- dim(unmethOxBS)[1]
nSpecimens <- dim(unmethOxBS)[2]

# Create container for the OxyBS results
MethOxy <- array(NA,dim=c(nCpGs,nSpecimens,3))
dimnames(MethOxy) <- list(
  rownames(methBS)[1:nCpGs],
  colnames(methBS)[1:nSpecimens], c("C","5mC","5hmC"))

# Process results (one array at a time, slow)
for(i in 1:nSpecimens){
MethOxy[,i,] <-fitOxBS(betaBS[,i],betaOxBS[,i],signalBS[,i],signalOxBS[,i])
}
```

```{r,echo=FALSE,eval=FALSE}
save(MethOxy,file="MethOxy.rds")
```


```{r,echo=FALSE,eval=TRUE}
load("MethOxy.rds")
```

Plot of the results (we have 4 replicates)

```{r,fig.align='center',fig.height=3,fig.width=9}

par(mfrow =c(1,3)) 
plot(density(MethOxy[,1,3]),main= "5-hmC using OxyBS",xlab="")
lines(density(MethOxy[,2,3]),col=2)
lines(density(MethOxy[,3,3]),col=3)
lines(density(MethOxy[,4,3]),col=4)

plot(density(MethOxy[,1,2]),main= "5-mC using OxyBS",ylim=c(0,5),xlab="")
lines(density(MethOxy[,2,2]),col=2)
lines(density(MethOxy[,3,2]),col=3)
lines(density(MethOxy[,4,2]),col=4)

plot(density(MethOxy[,1,1]),main= "5-C using OxyBS",ylim=c(0,5),xlab="")
lines(density(MethOxy[,2,1]),col=2)
lines(density(MethOxy[,3,1]),col=3)
lines(density(MethOxy[,4,1]),col=4)
```


## Comparison of 5-hmC estimates from different methods


```{r,message=FALSE,warning=FALSE,fig.align='center'}
library(GGally)
# data for replicate 1 is shown
df <- data.frame(x = as.numeric(results_exact$hmC[,1]),y=as.numeric(results_em$hmC[,1]),
                 z = as.numeric(MethOxy[,1,3]),w=as.numeric(hmC_naive[,1]))
ggpairs(df, title = "5-hmc estimates",  
  axisLabels = "show",columnLabels=c("Exact MLE","EM","OxyBS","Naive"))
```


```{r,message=FALSE,warning=FALSE,fig.align='center',fig.height=5}
library(ggplot2)
ggplot(df,aes(x=x,y=z)) + geom_point(alpha = 0.3) + xlab("Exact MLE") +
  ylab("OxyBS")

ggplot(df,aes(x=y,y=z)) + geom_point(alpha = 0.3) + xlab("EM") +
  ylab("OxyBS")

```
