---
title: "`MLML2R`: an R package for maximum likelihood estimates of DNA methylation and hydroxymethylation"
author: Samara F. Kiihl, Maria Tellez-Plaza
output:
  html_document:
    keep_md: yes
    toc: true
    number_sections: true
---


# Introduction


This document presents an example of the usage of the `MLML2R` package for R.

Install the R package using the following commands on the R console:

```{r,echo=TRUE,eval=FALSE}
install.packages("devtools")
devtools::install_github("samarafk/MLML2R")
library(MLML2R)
```


Proposed analyses of single-base profiling of either 5-hmC or 5-mC require combining data obtained using bisulfite conversion, oxidative bisulfite conversion or Tet-Assisted bisulfite conversion methods, but doing so naively produces inconsistent estimates of 5-mC or 5-hmC level [(Qu *et al.*, 2013)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3789553/). 


The function `MLML` provides maximum likelihood estimates (MLE) for 5-hmC and 5-mC levels using data from any combination of two of the methods: BS-seq, TAB-seq or oxBS-seq. The function also provides MLE when combining these three methods.

The algorithm implemented in the `MLML` function is based on the Expectation-Maximization (EM) algorithm proposed by [Qu *et al.* (2013)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3789553/). In addition, when only two methods are combined, our implementation is optimized, since we derived the constrained exact MLE in analytical form for 5-mC or 5-hmC levels, and the iterative EM algorithm is not needed. Our improved formulation can, thus, decrease analytic processing time and computational burden, common bottlenecks when processing single-base profiling data from thousands of samples.

Furthermore, our routine is flexible and can be used with both next generation sequencing and Infinium Methylation microarray data in the R-statistical language.



# Preparing dataset

We will use the dataset from [Johnson *et al.* (2016)](https://www.ncbi.nlm.nih.gov/pubmed/27886174), which consists of 30 DNA samples from glioblastoma tumors treated with oxBS-BS and hybridized to the Infinium 450K array.

The steps shown in this section follows the [vignette](https://kasperdanielhansen.github.io/genbioconductor/html/minfi.html) from `minfi` package.

## Getting publicly available data

We start with the steps to get the raw data from the GEO repository.
The dataset from [Johnson *et al.* (2016)](https://www.ncbi.nlm.nih.gov/pubmed/27886174) is available at GEO accession [GSE73895](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE73895).

The sample was divided into four BS and four oxBS replicates.
 	
Platform used: 	GPL13534	Illumina HumanMethylation450 BeadChip (HumanMethylation450_15017482)



This example has the following dependencies:

```{r,warning=FALSE,message=FALSE}
library(minfi)
library(GEOquery)
```

Use the following commands to install these packages in R:
```{r,echo=TRUE,eval=FALSE}
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("minfi", "GEOquery"))
```



```{r,echo=FALSE}
if(! file.exists("GSE73895/GSE73895_RAW.tar"))
{
getGEOSuppFiles("GSE73895")
untar("GSE73895/GSE73895_RAW.tar", exdir = "GSE73895/idat")
head(list.files("GSE73895/idat", pattern = "idat"))
}
```


```{r,echo=TRUE,eval=FALSE}
getGEOSuppFiles("GSE73895")
untar("GSE73895/GSE73895_RAW.tar", exdir = "GSE73895/idat")
head(list.files("GSE73895/idat", pattern = "idat"))
```


Decompress the compressed IDAT files:

```{r,message=FALSE,warning=FALSE}
idatFiles <- list.files("GSE73895/idat", pattern = "idat.gz$", full = TRUE)
sapply(idatFiles, gunzip, overwrite = TRUE)
```


Now we read the IDAT files in the directory:

```{r}
rgSet <- read.metharray.exp("GSE73895/idat")
rgSet
```

```{r}
pData(rgSet)
```

```{r}
sampleNames(rgSet)
```

The file names consists of a GEO identifier (the GSM part) followed by a standard IDAT naming convention with a 10 digit number which is an array identifier followed by an identifier of the form R01C01. This is because each array actually allows for the hybridization of 12 samples in a 6x2 arrangement. The 3999941120_R03C01 means row 3 and column 1 on chip 3999941120. 

We need to identify the samples from different methods: BS-conversion, oxBS-conversion.

```{r,echo=FALSE}
if (!file.exists("pD.rds"))
{
geoMat <- getGEO("GSE73895")
pD.all <- pData(geoMat[[1]])
pD <- pD.all[, c("title", "geo_accession", "characteristics_ch1.1", "characteristics_ch1.2","characteristics_ch1.3")]
save(pD,file="pD.rds")
}
```


```{r,eval=FALSE}
geoMat <- getGEO("GSE73895")
pD.all <- pData(geoMat[[1]])
pD <- pD.all[, c("title", "geo_accession", "characteristics_ch1", "characteristics_ch1.2","characteristics_ch1.3")]
pD
```

```{r,echo=FALSE}
load("pD.rds")
pD
```

```{r}
names(pD)[c(1,3,4,5)] <- c("method","gender","survival_months","age_years")
pD$gender <- sub("^gender: ", "", pD$gender)
pD$age_years <- as.numeric(sub("^subject age: ", "", pD$age_years))
pD$survival_months <- as.numeric(sapply(pD$survival_months, function(x) strsplit(as.character(x),":")[[1]][2]))
pD$method <- sapply(pD$method, function(x) strsplit(as.character(x),"_")[[1]][3])
```

We now need to merge this pheno data into the methylation data. The following are commands to make sure we have the same row identifier in both datasets before merging.

```{r}
sampleNames(rgSet) <- sapply(sampleNames(rgSet),function(x) strsplit(x,"_")[[1]][1])
rownames(pD) <- pD$geo_accession
pD <- pD[sampleNames(rgSet),]
pData(rgSet) <- as(pD,"DataFrame")
rgSet
```


## Preprocessing

We refer the reader to the `minfi` package tutorials for more preprocessing options.


We need to install the required package bellow:
```{r,eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("IlluminaHumanMethylation450kmanifest")
```

The `rgSet` object is a class called `RGChannelSet` which represents two color data with a green and a red channel. We will use, as input in the `MLML` funcion, a `MethylSet`, which contains the methylated and unmethylated signals. The most basic way to construct a `MethylSet` is to using the function `preprocessRaw` which uses the array design to match up the different probes and color channels to construct the methylated and unmethylated signals. Here we will use the `preprocessNoob` function, which does the preprocessing and returns a `MethylSet`.


Arrays were then normalized using the Noob/ssNoob preprocessing method for Infinium methylation microarrays.


From a `MethylSet` it is easy to compute Beta values, defined as:

Beta = Meth / (Meth + Unmeth + c)

The c constant is chosen to avoid dividing with small values. Illumina uses a default of c=100. The function `getBeta` from `minfi` package can be used to obtain the Beta values.


```{r,echo=TRUE,message=FALSE}
MSet.noob<- preprocessNoob(rgSet)
densityPlot(MSet.noob, sampGroups= pData(rgSet)$method,
main= sprintf('Beta values', nrow(MSet.noob)))
```


# Using the `MLML2R` package

After all the preprocessing procedures, we now can use the `MLML2R` package to obtain the maximum likelihood estimates for the 5-hmC and 5-mC levels.

Install the R package using the following commands on the R console:


```{r,eval=FALSE}
install.packages("devtools")
devtools::install_github("samarafk/MLML2R")
```


Prepare de input data:


```{r,echo=TRUE}
BS_index <- which(pData(rgSet)$method=="BS")
oxBS_index <- which(pData(rgSet)$method=="oxBS")


MethylatedBS <- getMeth(MSet.noob)[,BS_index]
UnMethylatedBS <- getUnmeth(MSet.noob)[,BS_index]

MethylatedOxBS <- getMeth(MSet.noob)[,oxBS_index]
UnMethylatedOxBS <- getUnmeth(MSet.noob)[,oxBS_index]
```



Getting the MLE estimates using EM-algorithm:

```{r}
library(MLML2R)
results_em <- MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, M = MethylatedOxBS,tol=0.0001)
```
```{r,fig.align='center',fig.height=3,fig.width=9}
par(mfrow =c(1,3)) 
densityPlot(results_em$hmC,main= "5-hmC using EM-algortihm")
densityPlot(results_em$mC,main= "5-mC using EM-algortihm")
densityPlot(results_em$C,main= "5-C using EM-algortihm")
```


Getting the constrained exact MLE estimates:

```{r}
library(MLML2R)
results_exact <- MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, M = MethylatedOxBS,exact=TRUE)
```

```{r,fig.align='center',fig.height=3,fig.width=9}
par(mfrow =c(1,3)) 
densityPlot(results_em$hmC,main= "5-hmC using constrained exact MLE")
densityPlot(results_em$mC,main= "5-mC using constrained exact MLE")
densityPlot(results_em$C,main= "5-C using constrained exact MLE")
```


Comparing the two methods:

```{r}
all.equal(results_exact$hmC,results_em$hmC)
cor(results_exact$hmC,results_em$hmC)

library(microbenchmark)
mbm = microbenchmark(
  EM = MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, 
            M = MethylatedOxBS,tol=0.0001),
  EXACT = MLML(T = MethylatedBS , U = UnMethylatedBS, L = UnMethylatedOxBS, 
               M = MethylatedOxBS,exact=TRUE)
)
mbm
```

# Other methods to obtain the estimates

## Naive estimates

The naive approach to obtain 5-hmC levels is $\beta_{BS} -  \beta_{OxBS}$. This approach results in negative values for the 5-hmC levels.

```{r,echo=TRUE}
beta_BS <- getBeta(MSet.noob)[,BS_index]
beta_OxBS <- getBeta(MSet.noob)[,oxBS_index]
hmC_naive <- beta_BS-beta_OxBS
C_naive <- 1-beta_BS
mC_naive <- beta_OxBS
```


```{r,fig.align='center',fig.height=3,fig.width=9}
par(mfrow =c(1,3)) 
densityPlot(hmC_naive,main= "5-hmC using naive method")
densityPlot(mC_naive,main= "5-mC using naive method")
densityPlot(C_naive,main= "5-C using naive method")
```


## `OxyBS` estimates

For the specific case where only ox-BS and BS data are available, `OxyBS` package from [Houseman *et al.* (2016)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4978924/) can be use to obtain estimates.



```{r,echo=TRUE,eval=FALSE}
library(OxyBS)

# Methylated signals from the BS and oxBS arrays
methBS <- MethylatedBS
methOxBS <- MethylatedOxBS
# Unmethylated signals from the BS and oxBS arrays
unmethBS <- UnMethylatedBS
unmethOxBS <- UnMethylatedOxBS

# Calculate Total Signals
signalBS <- methBS+unmethBS
signalOxBS <- methOxBS+unmethOxBS

# Calculate Beta Values
betaBS <- methBS/signalBS
betaOxBS <- methOxBS/signalOxBS

####################################################
# 4. Apply fitOxBS function to preprocessed values
####################################################

# Select the number of CpGs and Subjects to which the method will be applied 
nCpGs <- dim(unmethOxBS)[1]
nSpecimens <- dim(unmethOxBS)[2]

# Create container for the OxyBS results
MethOxy <- array(NA,dim=c(nCpGs,nSpecimens,3))
dimnames(MethOxy) <- list(
  rownames(methBS)[1:nCpGs],
  colnames(methBS)[1:nSpecimens], c("C","5mC","5hmC"))

# Process results (one array at a time, slow)
for(i in 1:nSpecimens){
MethOxy[,i,] <-fitOxBS(betaBS[,i],betaOxBS[,i],signalBS[,i],signalOxBS[,i])
}
```

```{r,echo=FALSE,eval=FALSE}
save(MethOxy,file="MethOxy.rds")
```




```{r,echo=FALSE,eval=TRUE}
load("MethOxy.rds")
```

Plot of the results (we have 4 replicates)

```{r,fig.align='center',fig.height=3,fig.width=9}

par(mfrow =c(1,3)) 
densityPlot(MethOxy[,,3]),main= "5-hmC using OxyBS",xlab="")
densityPlot(MethOxy[,,2]),main= "5-mC using OxyBS",xlab="")
densityPlot(MethOxy[,,1]),main= "5-C using OxyBS",xlab="")
```


## Comparison of 5-hmC estimates from different methods


```{r,message=FALSE,warning=FALSE,fig.align='center'}
library(GGally)
# data for replicate 1 is shown
df <- data.frame(x = as.numeric(results_exact$hmC[,1]),y=as.numeric(results_em$hmC[,1]),
                 z = as.numeric(MethOxy[,1,3]),w=as.numeric(hmC_naive[,1]))
ggpairs(df, title = "5-hmc estimates",  
  axisLabels = "show",columnLabels=c("Exact MLE","EM","OxyBS","Naive"))
```


```{r,message=FALSE,warning=FALSE,fig.align='center',fig.height=5}
library(ggplot2)
ggplot(df,aes(x=x,y=z)) + geom_point(alpha = 0.3) + xlab("Exact MLE") +
  ylab("OxyBS")

ggplot(df,aes(x=y,y=z)) + geom_point(alpha = 0.3) + xlab("EM") +
  ylab("OxyBS")

```


